Tutorial: Fitting log-Gaussian mixed-effects mixture models for keystroke data

### 2024-05-28 tutorial/tutorial.Rmd
minor fix
Submitted paper

This is a step-by-step tutorial for fitting mixed-effects mixture models with two log-Gaussian distributions [@roeser2021] using R [@R-base] and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2]. We will compare the mixture-model analysis to standard Gaussian and log-Gaussian mixed-effect models fitted using `brms` [@burkner2017brms; @R-brms_b].^[The R package `brms` provides a flexible framework to implement mixture models and many other types of probability models [@burkner2017brms; @R-brms_b]. `brms` has a `mixture` function to specify mixtures of various types of distributions. These could be Gaussian, skewed-Normal, shifted-Normal, ex-Gaussian etc. and combinations thereof. There is a large number of probability models for continuous data that are plausible candidates [for reaction time data see e.g. @matzke2009psychological]. In this tutorial we use Stan for mixture modelling to demonstrate the process and because the publication associated with this tutorial is fully using Stan instead of `brms`.]

In this guide we show how to fit a mixture model written in Stan to keystroke data. Then we demonstrate how the posterior of this model can be used to calculate differences between transition locations in the text. Also we show how to compare model using leave-one-out cross-validation. This tutorial is largely self-contained. To run the code below, the reader only needs to install the required packages. Data and the Stan programme can be loaded using the code below.

Three packages need to be installed before (see footnotes for instructions): (1) `rstan` to use R to interface with Stan for fitting Bayesian models^[Instructions on how to install `rstan` can be found [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for data wrangling and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` for fitting Bayesian models.^[Run `install.packages("brms")` for installation.]

@@

First we load a sample of keystroke data. Data are loaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/) using the URL as below. For simplicity we use a random subset of data from participants who completed an essay-writing task in the context of the ProWrite project.

```{r eval = F}
# Load data
data <- read_csv("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/data/sampledata.csv") 
```

```{r echo = F}
# Load data
data <- read_csv("data/sampledata.csv") 
```

We reduced the data to three variables: a participant identifier `participant`, the interkey intervals `iki`, and the transition location `location` (levels: within-word, before-word, before-sentence).

@@

We reduced the number of interkey intervals to `n` observations per participant and per location.

@@

We start with two standard single distribution mixed-effects models [as in @pinheiro2007linear; @baa08], i.e. models that assume that data come from a single underlying process that can be described as Gaussian or log-Gaussian. We will compare these models.

@@

We start with a Gaussian mixed-effects model. We describe the process that generates the $i^\text{th}$ interkey interval, where $i \in 1\dots N$ and $N$ is the total number of keystroke, as a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$. The mean can be decomposed into $\beta$ and $u_\text{participant[i]}$. $\beta$ will be allowed to take on a different value for each transition location associated with the $i^\text{th}$ keystroke. By-participant interkey intervals (ikis) are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant ikis, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.

$$
\begin{align}
\text{iki}_i \sim & \mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
\mu_i = & \beta_\text{location[i]} + u_\text{participant[i]}\\
u_\text{participant[i]} \sim & \mathcal{N}(0, \sigma_\text{p}^2) 
\end{align}
$$
This model can be described in R using the following formula wrapped in the `brms` formula function `bf`.

@@

@@

`brms` is using sensible regulating default priors for most model parameters which can be inspected using `get_prior` which needs the model formula and the data as input.

@@

We want to make sure that `(flat)` is replaced by more (i.e. weakly or regulating) informative priors [see @mcelreath2016statistical; @kruschke2010doing; @lambert2018student]. For example, we might consider a Student-*t* distribution as prior over the slope parameters (class `b`) which in our model are parameterised as the cell means of the three transition locations (levels: within word, before word, before sentence) with a mean of 350 msecs and a standard deviation of 75 msecs.

@@

Student-*t* distributions, in comparison to normal distributions, have a degrees of freedom parameter (first argument) in addition to the mean and the standard deviation (second and third parameter, respectively). The degrees of freedom control the tails of the distribution with smaller values assigning more probability mass to the tails; the shape of the Student-*t* distribution is moving closer to a normal distribution when the degrees of freedom approach $\infty$. This is illustrated in Figure \@ref(fig:prior) for the mean and the standard deviation of the prior above.

@@

The choice of priors here is weakly informative. The intention is to use knowledge that we generally have about how long interkey intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values. Even if we know very little about the duration of interkey intervals, we do know that they shouldn't be longer than a day, or event an hour, or several minutes, because such long intervals are unlikely to reflect active writing.

@@

This model is largely identical to the Gaussian mixed-effect model in the previous section, except instead of a normal distribution this model assumes the ikis come from a single underlying process that follows a log-normal distribution. Assuming a log-normal distribution is useful as ikis and other response times are zero-bound (interkey intervals cannot be shorter than 0). A log-normal distribution forces responses to be positive. Also the log scale is not linear, meaning the distance between 100 msecs and 110 msecs is not the same as the distance between 500 and 510 msecs; instead the distance between adjacent values is reduced for larger values while the distances between adjacent values on the lower end is increased. For example, for a log with the base 10, the distance between 10 msecs and 100 msecs is identical to the distance between 100 msecs and 1000 msecs.

$$
\begin{align}
\text{iki}_i \sim & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
\mu_i = & \beta_\text{location[i]} + u_\text{participant[i]}\\
u_\text{participant[i]} \sim & \mathcal{N}(0, \sigma_\text{p}^2) 
\end{align}
$$
For the model formula we use the distribution family `lognormal`.

@@

The default priors and the priors that we need to specity are now no longer on the msecs scale but on the log scale.

@@

@@

The function `fixef` provide a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.

@@

The posterior estimates obtained from the log-Gaussian model are radically different (larger) from the posterior estimates of the Gaussian model which is likely due. This difference is likely due to the missing lower bound at zero in the Gaussian model as we as its inability to give less emphasis to values in the right tail.

@@

To evaluate the out-of-sample predictive performance of the models we can use leave-one-out cross-validation via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical]. Cross-validation is a alternative to more conventional model-comparison metrics such as $R^2$ as it penalises models with more parameters to prevent overfitting [see @mcelreath2016statistical; @lambert2018student]. Overfitting happens when a model is capturing nuances of a sample, rather than patterns one would expect to repeat in new samples.

@@

The function `loo` can be used to access the expected log-predictive density which uses probability calculations to approximate LOO-IC via Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical]. LOO can be approximated with the log posterior predictive densities as a measure of how likely a data point is, given the parameter estimates. This is shown for each data point in Figure \@ref(fig:loglik). Extremely small values indicate that an observation is very unlike under the model's parameter values and we are unlike to observe them in a new sample.

@@

If we sum over those means we obtain the $\widehat{elpd}$, the expected log predictive density, indicated as `elpd_loo`. $p_\text{loo}$ (`p_loo`) is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters). `looic`, or LOO-IC, corresponds to $-2 \times ($`elpd_loo`$-$`p_loo`$)$ (for the deviance scale). To illustrate these, we can calculate them manually

@@

The Pareto *k* diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model.

The difference in `elpd_loo` for both models can be obtained using `loo` and is shown in `elpd_diff` ($\Delta\widehat{elpd}$) with standard error shown in `se_diff`.

@@

The model comparison shows that the model `fit_lgaus` (the model with the log-Gaussian likelihood) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors better than `fit_gaus`. This can be calculated by `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])` by normalising the difference over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, which returns the *z* or *t* value of the difference between models [@sivula2020uncertainty]. A difference of `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors is substantial. By comparison, a *t*-value of more extreme than $\mid2\mid$ corresponds to a difference that is significant at a $\alpha$-level of 0.05 (i.e. $p<0.05$) [e.g. @baa08book].

Aside, unlike $R^2$, `elpd_loo` and `looic` have no direct interpretation. This is because the size of `elpd_loo` increases as a function of the number of observations. Generally you can think of `elpd_loo` and `looic` as how many possible futures the model can rule out. Important is the difference between models.

In contrast, `loo_R2` provides a Bayesian $R^2$ that measures the variance in the data that the model can capture. This quantity is not ideal for model comparisons because, similar to the frequentist $R^2$, it is subject to overfitting: a model with more parameters can never have a smaller $R^2$, and $R^2$ will innevitabily increase when we add more parameters to a model. A larger $R^2$ means that a model captures the data well but might make predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; also chapter 6 in @mcelreath2016statistical]. The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.

@@

The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$). In the model below, the latter two are estimated separately for each level of the categorical prediction `location` (i.e. transition location) for the $i^\text{th}$ observation.

@@

Model takes into account two sources of random error: (1) each participant has an individual fluent typing speed that differs across levels of the categorical predictor and is distribution $u_\text{participant} \sim \mathcal{N}(0, \sigma^2_p)$; (2) each participant has in individual hesitation probability $\theta_\text{participant}$ that differs across levels of the categorical predictor.

The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel). The illustrate shows the dissociation of two underlying processes that the model is estimating on the basis of the data. There are two distributions of which the left one in grey is the distribution of fluent interkey intervals; the right distribution in green is wider and captures longer interkey intervals (with some overlap between fluent and disfluent interkey intervals).

@@

ggplot(data_mm, aes(x)) +
  scale_x_continuous(labels = exp, breaks = log(c(0, 10, 100, 1000, 10000)), limits = log(c(30, 50000))) +
  scale_y_continuous(limits = c(0, .7)) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#000000",
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#009E73",
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#000000",
                show.legend = F, alpha = .25, 
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#009E73",
                show.legend = F, alpha = .25, 
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  annotate("label", x=4,y=.6, label="beta*':'~'Typing speed'", size = 4, parse = T) +
  annotate("label", x=7.25,y=.35, label="atop(theta*':'~'Hesitation probability', '\"Pausing\"')", parse = T, size = 4) +
  annotate("label", x=9,y=.15, parse = T, label="delta*':'~'\"Slowdown\"'", size = 4) +
  geom_segment(x = alpha, y = 0.05, xend = alpha2, yend = 0.05, 
               colour = "blue", alpha = .5,
               arrow = arrow(length = unit(0.2, "cm"), 
                             ends = "both", 
                             type = "closed")) +
  geom_segment(x = alpha, y = 0, xend = alpha, yend = Inf, colour = "darkred", alpha = .5) +
  geom_segment(x = alpha2, y = 0, xend = alpha2, yend = Inf, colour = "darkred", alpha = .5) +
  geom_curve(x = 4, y = .57, xend = alpha - .75, yend = 0.1, 
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 7.5, y = .3, xend = alpha2 + .85, yend = 0.1,
             curvature=-.3,
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 9, y = .125, xend = 5.75, yend = 0.0255,
             curvature=-.3,
   arrow = arrow(length = unit(0.03, "npc"))) +
  labs(x = "iki [in msecs]", 
       y = "Density",
       caption = "see Roeser et al. (2021)")

@@

Importantly Figure \@ref(fig:mixturemodel) illustrates the average typing speed of fluent interkey intervals captured by the $\beta$ parameter; the distance between fluent transitions and disfluent transitions captured by $\delta$ (i.e. the distance between the grey and the green distribution); the hesitation probability or the "pausing frequency" which is the height of the green distribution captured by $\theta$. Most interkey intervals are, typically, fluent and hence part of the grey distribution. The frequency of the minority of disfluent interkey intervals is captured by the height of the green distribution. As this model assumes a finite mixture of two distributions, the probability of fluent interkey intervals is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent interkey intervals is $.6$. If the pausing probability decreases to $.3$ the proportion of fluent interkey intervals had increased to $.7$ etc.

@@

Participants and conditions are expected to represent indices in the Stan code. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before sentence = 1, before word = 2, within word = 3). The returned values are indices for the parameters in the model: For example, for the location identifier, `delta[1]` is the population estimate for the slowdown for hesitant interkey intervals at before sentence transition locations, `delta[2]` for before word locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.

@@

In the below code, the names on the left side of the arrow must correspond to the names expected in the Stan code (above); names on the right side do not. We assign the participant and location identifiers created above and their maximum values `nS` and `K`, respectively. The interkey interval data `iki` are assigned to `y` and the total number of observations is assigned to `N` (number of rows in the data `nrow(data`).

@@

The Stan code can be downloaded from [GitHub](https://github.com/jensroes/prowrite-mixture-models/blob/main/tutorial/stan/molg.stan) using `readLines` and saved locally using `writeLines` as "molg.stan" (mixture of log-Gaussians) in the directory "stan".

```{r eval = F}
# Download Stan code
file <- readLines("https://raw.githubusercontent.com/jensroes/prowrite-mixture-models/main/tutorial/stan/molg.stan")
# Save Stan code in "stan" directory
writeLines(file, "stan/molg.stan")
```

@@

This code fits the posterior of a categorical predictor with any number of levels. In other words, the posterior that the model returns can be used to calculate simple differences, main effects and interactions of factorial designs. We demonstrate below how to calculate a simple difference. The same logic can be used to calculate main effects and interactions. Also, participant estimates are calculated for both mixture components and the hesitation probability by location. For including more predictors, or removing, e.g., random error terms, or adjusting priors, the Stan code can be changed by opening it in R or a text editor. The Stan code used is largely based on @sorensen2016bayesian and @vasishth2017. @sorensen2016bayesian presents a detailed tutorial on how to write models in Stan [see also @lambert2018student].

The model is fitting a number of parameters of which not all are interesting for inference. To reduce the size of the model object that contains the posterior we can determine which parameters should be saved and disregard all parameters.

@@

```{r eval = F}
# Fit model
fit_molg <- sampling(molg, 
                     data = data_list,
                     iter = iterations,
                     warmup = warmup,
                     chains = nchains, 
                     cores = ncores,
                     pars = pars, # Parameters to keep.
                     save_warmup = FALSE, # Don't save the warmup samples.
                     seed = 365, 
                     # Improve sampling efficiency
                     control = list(adapt_delta = 0.94,  # default: 0.9
                                    max_treedepth = 12)) # default: 10
```

@@

Running this model will take a while to complete sampling depending on your hardware specifications. The time it took my laptop to complete this job can me viewed using `get_elapsed_time`. It is worth to not use all cores to run this model or to use a dedicated high performance machine.

@@

Inferred posterior estimates shown as average and 95% probability interval can be extracted using the `print` function for each transition location and the three parameter values of interest. `prob` is the $\theta$ parameter on the probability scale rather than on the logit scale. $\beta$ and $\delta$ are on the log scale.

```{r}
print(fit_molg, 
      pars = c("beta", "delta", "prob"), 
      probs = c(.025, .975))
```

@@

Before we interpret these result we need to ensure that the model reached convergence. Convergence can be established using two simple techniques. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain in different colours. We will look at the population-level parameters saved in `pars`.

@@

As described above we used leave-one-out cross-validation to compare the mixture of log-Gaussians model to the single distribution Gaussian and log-Gaussian models. This time we need to use the `loo_compare` function because we had to extract the LOO statistics.

>>>>

This is a step-by-step tutorial for fitting mixed-effects mixture models with two log-Gaussian distributions [@peel2000finite] using R [@R-base] and the `rstan` package to interface with the probabilistic programming language Stan [@carpenter2016stan; @hoffman2014no; @rstan; @rstan2]. We will compare the mixture-model analysis to standard Gaussian and log-Gaussian mixed-effect models fitted using `brms` [@burkner2017brms; @R-brms_b].^[The R package `brms` provides a flexible framework to implement mixture models and many other types of probability models [@burkner2017brms; @R-brms_b]. `brms` has a `mixture` function to specify mixtures of various types of distributions. These could be Gaussian, skewed-Normal, shifted-Normal, ex-Gaussian etc. and combinations thereof. There is a large number of probability models for continuous data that are plausible candidates [for reaction time data see e.g. @matzke2009psychological]. In this tutorial we use Stan for mixture modelling to demonstrate the process and because the publication associated with this tutorial is fully using Stan instead of `brms`.]

In this tutorial we show how to fit a mixture model written in Stan to keystroke data [similar to @roeser2021] and how to compare this to model mainstream mixed-effects models using leave-one-out cross-validation. Then we demonstrate how the posterior of this model can be used to calculate differences between transition locations in the text. This tutorial is largely self-contained. To run the code below, the reader only needs to install the required packages. Data and the Stan programme can be loaded using the code below.

Three packages need to be installed before (see footnotes for instructions): (1) `rstan` to interface with Stan^[Instructions on how to install `rstan` can be found [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).]; (2) `tidyverse` for general data wrangling and visualisation.^[Run `install.packages("tidyverse")` for installation.]; (3) `brms` to for Bayesian models without directly working in Stan.^[Run `install.packages("brms")` for installation.]

@@

First we load a sample of keystroke data. Data are loaded from our [OSF](https://osf.io/z65dw/) repository using the URL as below. For simplicity we use a random subset of data from participants who completed an essay-writing task in the context of the ProWrite project [e.g. @dux2022automating].

```{r eval = F}
# Load data
data <- read_csv("https://osf.io/tjms9/download")
```

```{r echo = F}
# Load data
data <- read_csv("https://osf.io/tjms9/download")
```

```{r echo = F}
n_ppts <- data %>% pull(participant) %>% unique() %>% length()
ns <- count(data, participant, location) %>% pull(n) %>% unique()
```

We reduced the data to `r n_ppts` participants and three variables: a participant identifier `participant`, the interkey intervals `iki`, and the transition location `location` (levels: within word, before word, before sentence).

@@

We reduced the number of interkey intervals to `r ns` observations per participant and per transition location by randomly sampling interkey intervals. See for example:

@@

We start with standard single distribution mixed-effects models [as in e.g. @baa08; @pinheiro2007linear], that the data come from a single underlying process that, the first model, can be described as Gaussian and log-Gaussian for the other single distribution model. We compare these models below.

@@

We start with a Gaussian mixed-effects model. The process that generates the $i^\text{th}$ interkey interval (iki below), where $i \in 1\dots N$ and $N$ is the total number of keystroke, is a single Gaussian (normal) distribution $\mathcal{N}()$ which is characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$. This model can be expressed mathematically using the following equation:

$$
\begin{align}
\text{iki}_i \sim & \mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
\mu_i = & \beta_\text{location[i]} + u_\text{participant[i]}\\
u_\text{participant[i]} \sim & \mathcal{N}(0, \sigma_\text{p}^2) 
\end{align}
$$
The mean can be decomposed into $\beta$ and $u_\text{participant[i]}$. $\beta$ was allowed to take on a different value for each transition location associated with the $i^\text{th}$ interkey interval. By-participant interkey intervals are assumed to vary around the average which is achieved by placing a normal distribution on the difference between average and by-participant interkey intervals, indicated as $u_\text{participant[i]}$, with a mean of 0 and a standard deviation $\sigma_\text{p}^2$.

This model can be described in R using the following formula wrapped in the `brms` formula function `bf`. The `family` argument can be used to specify distributions where `gaussian` is the default can could be omitted below.

@@

The syntax used in `brms` is generally similar to `lme4` syntax that many readers will be familiar with. Like in `lm`, `glm`, `glmer`, `lmer` and other modelling functions `0 + ` can be used to drop the intercept parameter which means that the model is parametrised to return cell means and standard errors for each level of `location`.

@@

As we are modelling data in the Bayesian framework, we need to define priors before model fit. `brms` is using sensible regulating default priors for most model parameters which can be inspected using `get_prior` which needs the model formula defined above and the data as input.

@@

We want to make sure that `(flat)` is replaced by more (i.e. weakly or regulating) informative priors [see @kruschke2010doing; @lambert2018student; @mcelreath2016statistical]. For example, we might consider a Student-*t* distribution as prior over the slope parameters (class `b`) which in our model are parameterised as the cell means of the three transition locations (levels: within word, before word, before sentence) with a mean of 350 msecs and a standard deviation of 75 msecs.

@@

Student-*t* distributions, in comparison to normal distributions, have a degrees of freedom parameter (first argument) in addition to the mean and the standard deviation (second and third parameter, respectively). The degrees of freedom control the thickness of the tails of the distribution with smaller values assigning more probability mass to the tails; the shape of the Student-*t* distribution is moving closer to a normal distribution when the degrees of freedom approach $\infty$. This is illustrated in Figure \@ref(fig:prior) for the mean and the standard deviation of the prior above.

@@

These priors are weakly informative. The intention is to use general knowledge that we have about how long interkey intervals usually are to point the model towards the right area in the parameter space by assigning low probabilities to unlikely or even implausible values. Even if we know very little about the duration of interkey intervals, we do know that they shouldn't be longer than a day, or an hour, or even several minutes, because such long intervals are too extreme to reflect active writing.

@@

This model is largely identical to the Gaussian mixed-effect model in the previous section, except instead of a normal distribution this model assumes the interkey intervals come from a single underlying process that follows a log-normal distribution indicated as $\text{log}\mathcal{N}()$.

$$
\begin{align}
\text{iki}_i \sim & \text{log}\mathcal{N}(\mu_i, \sigma_\text{e}^2)\\
\mu_i = & \beta_\text{location[i]} + u_\text{participant[i]}\\
u_\text{participant[i]} \sim & \mathcal{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

Assuming a log-normal distribution is useful as interkey intervals and other response times are zero-bound (interkey intervals cannot be shorter than 0). A log-normal distribution forces responses to be positive. Also the log scale is not linear, meaning the distance between 100 msecs and 110 msecs is not the same as the distance between 500 and 510 msecs; instead the distance between adjacent values is reduced for larger values while the distances between adjacent values on the lower end is increased. For example, for a logarithm with the base 10, the distance between 10 msecs and 100 msecs is identical to the distance between 100 msecs and 1000 msecs.

For the model formula we need to change the distribution family from `gaussian` (default) to `lognormal`.

@@

The default priors and the priors that we need to specify are now no longer on the msecs scale but on the log scale.

@@

Here is again an example for different degrees of freedom for the prior above.

@@

The `fixef` function provides a summary of the inferred posterior estimates as average and 95% probability interval by transition location in log msecs.

@@

The posterior estimates obtained from the log-Gaussian model are radically different from the posterior estimates of the Gaussian model. This difference is likely due to the missing lower bound at zero in the Gaussian model as we as its inability to de-emphasize values in the right tail.

@@

To evaluate the out-of-sample predictive performance of the models we can use leave-one-out cross-validation via Pareto smoothed importance-sampling [@vehtari2015pareto; @vehtari2017practical]. Cross-validation is an alternative to more conventional model-comparison metrics such as $R^2$ as it penalises models with more parameters to prevent overfitting [see @lambert2018student; @mcelreath2016statistical]. Overfitting happens when a model captures nuances of a sample, rather than those patterns that one would expect to observe in new samples.

@@

The function `loo` can be used to access the expected log-predictive density which uses probability calculations to approximate the LOO-IC via Pareto smoothed importance sampling [@vehtari2015pareto; @vehtari2017practical]. LOO can be approximated with the log posterior predictive densities as a measure of how likely a data point is, given the parameter estimates. This is shown in Figure \@ref(fig:loglik) for each data point in the sample. Extremely small values indicate that an observation is very unlike to repeat in a new sample under the given model and the estimated posterior parameter values.

@@

If we sum over the means shown in Figure \@ref(fig:loglik), we obtain $\widehat{elpd}$, the expected log predictive density, indicated as `elpd_loo`. $p_\text{loo}$ (`p_loo`) is the sum of the variances and indicates the flexibility of the model fit (also used as the effective number of parameters). `looic`, or LOO-IC, corresponds to $-2 \times ($`elpd_loo`$-$`p_loo`$)$ (for deviance scale). To illustrate these, we can calculate them manually

@@

The Pareto *k* diagnostic indicates that there are some observations that are very unusual under the single distribution Gaussian model and the obtained parameter estimates.

The difference in `elpd_loo` for the two single distribution models is returned by `loo` and is shown in the `elpd_diff` column ($\Delta\widehat{elpd}$) with standard error shown in the `se_diff` column.

@@

The model comparison shows that the model `fit_lgaus` (the log-Gaussian model) is approximately `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors better than `fit_gaus`. This can be calculated by `abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2])` by normalising the difference over its standard error $\mid\frac{\Delta\widehat{elpd}}{\text{SE}}\mid$, which returns the difference between models expressed in standard errors [@sivula2020uncertainty]. A difference of `r round(abs(loo_fit$diffs[2,1] / loo_fit$diffs[2,2]),0)` standard errors is substantial. By comparison, a *t*-value more extreme than $\mid2\mid$ corresponds to an effect that is significant at an $\alpha$-level of 0.05 (i.e. $p<0.05$) [e.g. @baa08book].

Aside, unlike $R^2$, `elpd_loo` and `looic` have no direct interpretation. This is because the size of `elpd_loo` increases as a function of the number of observations. Generally `elpd_loo` and `looic` can be understood as a quantification of how many possible futures the model can rule out. However, what is of importance is the difference between models.

If a measure of variance explained is desired, `loo_R2` provides a Bayesian $R^2$. This quantity is not ideal for model comparisons because, similar to the frequentist $R^2$, it is subject to overfitting: $R^2$ will inevitabily increase when more parameters are added to the model. A large $R^2$ means that a model captures the data well but might mean that the model makes predictions that do not generalise out-of-sample [see chapter 11 in @gelman2020regression; also chapter 6 in @mcelreath2016statistical]. The is an example for Bayesian $R^2$ for the single-distribution log-Gaussian model.

@@

The mixture model captures the writing process as a combination of three parameters: the average typing speed or fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$). In the model below, the latter two are estimated separately for each level of the categorical prediction `location` (i.e. transition location) for the $i^\text{th}$ observation.

@@

This model takes into account two sources of random error: (1) each participant has an individual typing speed modelled as deviations from the average $u_\text{participant} \sim \mathcal{N}(0, \sigma^2_p)$ as above; (2) each participant has in individual hesitation probability $\theta_\text{participant}$ that differs across levels of the categorical predictor `location`.

The parameters of this equation are visualised in Figure \@ref(fig:mixturemodel). This illustration shows the dissociation of two underlying processes that the model is estimating on the basis of the data. There are two distributions of which the left one in grey is the distribution of fluent interkey intervals; the right distribution in green is wider and captures longer interkey intervals (with some overlap between fluent and disfluent interkey intervals).

@@

ggplot(data_mm, aes(x)) +
  scale_x_continuous(labels = exp, breaks = log(c(0, 10, 100, 1000, 10000)), limits = log(c(30, 50000))) +
  scale_y_continuous(limits = c(0, .7)) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#000000",
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "#009E73",
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#000000",
                show.legend = F, alpha = .25, 
                args = list(alpha[1], sigma[1], lam = 1 - theta[1])) +
  stat_function(geom = "area", fun = plot_mix_comps, fill = "#009E73",
                show.legend = F, alpha = .25, 
                args = list(alpha2[1], sigmap[1], lam = theta[1])) +
  annotate("label", x=4,y=.6, label="beta*':'~'Typing speed'", size = 4, parse = T) +
  annotate("label", x=7.25,y=.35, label="atop(theta*':'~'Hesitation probability', '\"Pausing\"')", parse = T, size = 4) +
  annotate("label", x=9,y=.15, parse = T, label="delta*':'~'\"Slowdown\"'", size = 4) +
  geom_segment(x = alpha, y = 0.05, xend = alpha2, yend = 0.05, 
               colour = "blue", alpha = .5,
               arrow = arrow(length = unit(0.2, "cm"), 
                             ends = "both", 
                             type = "closed")) +
  geom_segment(x = alpha, y = 0, xend = alpha, yend = Inf, colour = "darkred", alpha = .5) +
  geom_segment(x = alpha2, y = 0, xend = alpha2, yend = Inf, colour = "darkred", alpha = .5) +
  geom_curve(x = 4, y = .57, xend = alpha - .75, yend = 0.1, 
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 7.5, y = .3, xend = alpha2 + .85, yend = 0.1,
             curvature=-.3,
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(x = 9, y = .125, xend = 5.75, yend = 0.0255,
             curvature=-.3,
   arrow = arrow(length = unit(0.03, "npc"))) +
  labs(x = "iki [in msecs]", 
       y = "Density")

@@

Importantly Figure \@ref(fig:mixturemodel) illustrates the average typing speed of fluent interkey intervals captured by the $\beta$ parameter; the distance between fluent transitions and disfluent transitions captured by $\delta$ (i.e. the distance between the grey and the green distribution); the hesitation probability or the "pausing frequency" which is the height of the green distribution captured by $\theta$.

Most interkey intervals are, typically, fluent and hence part of the grey distribution. The frequency of the minority of disfluent interkey intervals is captured by the height of the green distribution. As this model assumes a finite mixture of two distributions, the probability of fluent interkey intervals is the inverse $1-\theta$; in other words, a pausing probability of $\theta=.4$ means that the probability of fluent interkey intervals is $.6$. If the pausing probability decreases to $.3$ the proportion of fluent interkey intervals had increased to $.7$.

@@

Participants and conditions are expected to represent indices in the Stan code. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before sentence = 1, before word = 2, within word = 3). The returned values are indices for the parameters in the model: For example, for the location identifier, `delta[1]` is the population estimate for the slowdown for hesitant interkey intervals at before sentence locations, `delta[2]` for before word locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transitions, respectively.

@@

In the below code, the names on the left side of the arrow must correspond to the names expected in the Stan code (above); names on the right side do not. We assigned the participant and location identifiers created above and their maximum values `nS` and `K`, respectively. The interkey interval data `iki` were assigned to `y` and the total number of observations (number of rows in the data `nrow(data)`) was assigned to `N`.

@@

The Stan code can be downloaded from our [OSF](https://osf.io/z65dw/) repository using `readLines` and saved locally using `writeLines` as "molg.stan" (mixture of log-Gaussians) in the directory "stan".

```{r eval = F}
# Download Stan code
file <- readLines("https://osf.io/r7aqx/download")
# Save Stan code in "stan" directory
writeLines(file, "stan/molg.stan")
```

@@

This code can fit the posterior of a categorical predictor with any number of levels. In other words, the posterior that the model returns can be used to calculate simple differences, main effects and interactions of factorial designs. We demonstrate below how to calculate a simple difference. The same logic can be used to calculate main effects and interactions. For more predictors, or removing or adding, e.g., random error terms, or adjusting priors, the Stan code can be opened and changed in R or any text editor. The Stan code used is largely based on @vasishth2017. @sorensen2016bayesian presents a detailed tutorial on how to write models in Stan [see also @lambert2018student].

This model estimates a number of parameters of which not all are interesting for inference. To reduce the physical size of the model object that contains the posterior we can determine which parameters we want to save.

@@

```{r eval = F}
# Fit model
fit_molg <- sampling(molg, 
                     data = data_list,
                     iter = iterations,
                     warmup = warmup,
                     chains = nchains, 
                     cores = ncores,
                     pars = pars, # Parameters to keep.
                     save_warmup = FALSE, # Don't save the warmup samples.
                     seed = 365, 
                     # Improve sampling efficiency
                     control = list(adapt_delta = 0.96,  # default: 0.9
                                    max_treedepth = 14)) # default: 10
```

@@

Running this model will take a moment to complete sampling depending on your hardware specifications. The time it took my laptop to complete this job can be viewed using `get_elapsed_time`.

@@

Inferred posterior estimates shown as averages and 95% probability intervals can be extracted using the `print` function for each transition location and the three parameter values of interest. `prob` is the $\theta$ parameter on the probability scale rather than on the logit scale. $\beta$ and $\delta$ are on the log scale.

```{r}
print(fit_molg, 
      pars = c("beta", "delta", "prob", "theta"), 
      probs = c(.025, .975))
```

@@

Before we interpret these results we need to ensure that the model reached convergence. Convergence can be established using two simple methods. First, we can inspect the chains in trace plots which show the parameter estimate across iterations (here after warmup) for each chain in different colours. We will look at the population-level parameters saved in `pars`.

@@

In "Model comparison I" we used `loo` to compare two models directly. This time we need the `loo_compare` function because we had to extract the LOO statistics for comparison.
